\textbf{systemfit}\\ Simultaneous Equation Systems in R

### 2006-06-12 paper/branches/useR2006/Slides/systemfit_slides.tex
useR! 2006 Slides: added notes; some other improvements; removed slide about advantages
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@675 591db2bd-e0a0-4d22-8c19-2192dbb7da31

\begin{document}
\frame{ \titlepage }
\note{
\begin{itemize}
\item Welcome!
\item authors
\item package systemfit: Estimating simultaneous equation systems in R
\end{itemize}
}

@@

% ----------------- Motivation ----------------------------
\subsection{Motivation}
\frame{
   \frametitle{Motivation}
\begin{itemize}
\item many theoretical models consist of more than one equation
   \begin{itemize}
   \item contemporaneous correlation of disturbance terms (likely)
   \item simultaneous estimation of all equations as
      °Seemingly Unrelated Regression° (SUR) leads to efficient results
   \end{itemize}
\item theoretically derived cross-equation parameter restrictions\\
   \begin{itemize}
   \item simultaneous estimation of all equations required
   \end{itemize}
\item endogeneity of some variables\\
   \begin{itemize}
   \item estimation using °Two-Stage Least Squares° (2SLS)
      or °Three-Stage Least Squares° (3SLS) required
   \end{itemize}
\end{itemize}
}
\note{
\begin{itemize}
\item ?
\end{itemize}
}

% ----------------- Outline ----------------------------
\subsection{Outline}
\frame{
   \frametitle{Outline}
\begin{itemize}
\item Introduction
\item Features of systemfit
\item Examples
\item Advantages
\item Plans for the Future
\end{itemize}
}
\note{
\begin{itemize}
\item ?
\end{itemize}
}

@@

% ----------------- Estimation Methods ----------------------------
\subsection{Methods}
\frame{
   \frametitle{Estimation Methods}
\begin{itemize}
\item Ordinary Least Squares (OLS)
\item Two-Stage Least Squares (2SLS)
\item Seemingly Unrelated Regression (SUR)
\item Three-Stage Least Squares (3SLS)
\item \ldots
\end{itemize}
}
\note{
\begin{itemize}
\item 2SLS often used for single-equations
\end{itemize}
}

% ----------------- Estimation Control ----------------------------
\subsection{Estimation Control}
\frame{
   \frametitle{Estimation Control}
\begin{itemize}
\item imposition of linear restrictions
\item instrumental variables
\item formulas for the residual covariance matrix
\item formulas for 3SLS estimation
\item degrees of freedom for $t$ tests
\item homogenous residual variance in OLS/2SLS estimations
\end{itemize}
}
\note{
\begin{itemize}
\item imposition of linear restrictions by two different methods
\item instrumental variables: the same for all equations or different for each equation
\item several different formulas to calculate the residual covariance matrix
\item several different formulas for the 3SLS estimation
\item $t$ tests: degrees of freedom of the whole system
   or degrees of freedom of the single equation
\item OLS/2SLS estimation: different $\sigma^2$s for each single equation
   or the same $\sigma^2$ for all equations
\end{itemize}
}

% ----------------- Other Tools ----------------------------
\subsection{Other Tools}
\frame{
   \frametitle{Other Tools}
\begin{itemize}
\item wrapper function for (classical) panel-like data in long format:
   \texttt{systemfitClassic}
\item testing linear hypotheses using the F-, Wald-, and LR-statistic
\item Hausman test for the consistency of the 3SLS estimator
\end{itemize}
}
\note{
\begin{itemize}
\item ?
\end{itemize}
}

@@

% ----------------- Commands ----------------------------
\subsection{Commands}
\begin{SaveVerbatim}{specSystem}
eqDemand <- consump ~ price + income
eqSupply <- consump ~ price + farmPrice + trend
system <- list(demand=eqDemand, supply=eqSupply)
\end{SaveVerbatim}
% \end{Verbatim}
\begin{SaveVerbatim}{estSystem}
fitsur <- systemfit("SUR", system, data=Kmenta)
\end{SaveVerbatim}
% \end{Verbatim}
\begin{SaveVerbatim}{printResults}
summary( fitsur )
\end{SaveVerbatim}
% \end{Verbatim}
\frame{
   \frametitle{Example: Commands}
\begin{itemize}
\item from Kmenta (1986): Elements of Econometrics, p.~685
\item specification of the equation system:\\[2mm]
   {\small \BUseVerbatim{specSystem}}
\item estimation using method °SUR°:
   {\small \BUseVerbatim{estSystem}}
\item printing summary results:\\
   {\small \BUseVerbatim{printResults}}
\end{itemize}
}
\note{
\begin{itemize}
\item ?
\end{itemize}
}

@@

% ============== Advantages ==============================
\section{Advantages}
\frame{
   \frametitle{Advantages}
\begin{itemize}
\item all estimation procedures are publicly available
\item estimation algorithms can be easily modified
\item many estimation details can be controlled by the user
\item novices can safely ignore these estimation details
\item reliability of \texttt{systemfit} has been demonstrated
\end{itemize}
}
\note{
\begin{itemize}
\item all estimation procedures are publicly available in the source code
\item estimation algorithms can be easily modified to meet specific
requirements
\item the (advanced) user can control many estimation details generally
not available in other software packages
\item novices can safely ignore these estimation details
because they have reasonable defaults
\item reliability of \texttt{systemfit} has been demonstrated by comparing its estimation
results with results published in the literature
\end{itemize}
}

@@

% ----------------- General Plans for the Future ----------------------------
\subsection{General}
\frame{
   \frametitle{Plans for the Future}
\begin{itemize}
\item estimation with unbalanced data sets
\item estimation methods: LIML, FIML,  and GMM
\item fitting equation systems with serially correlated and
heteroscedastic disturbances
\item spatial econometric methods
\item improving the function \texttt{nlsystemfit}
to estimate systems of non-linear estimations
\end{itemize}
}
\note{
\begin{itemize}
\item ?
\end{itemize}
}

>>>>

\begin{document}
\frame{ \titlepage }
\note{
\begin{itemize}
\item Welcome to my presentation about the R package systemfit.
\item This package is written by me, Arne Henningsen,
   and by my co-author Jeff Hamann.
\item The package can used to estimate simultaneous equation systems in R.
\end{itemize}
}

@@

% ----------------- Motivation ----------------------------
\subsection{Motivation}
\frame{
   \frametitle{Motivation}
Many theoretical models consist of more than one equation
   \begin{itemize}
   \item contemporaneous correlation of disturbance terms (likely)
   \item simultaneous estimation of all equations as
      °Seemingly Unrelated Regression° (SUR) leads to efficient results
   \end{itemize}
\medskip
Theoretically derived cross-equation parameter restrictions\\
   \begin{itemize}
   \item simultaneous estimation of all equations required
   \end{itemize}
\medskip
Endogeneity of some variables
   \begin{itemize}
   \item estimation using °Two-Stage Least Squares° (2SLS)
      or °Three-Stage Least Squares° (3SLS) required
   \end{itemize}
\bigskip
$\Rightarrow$ All these models can be estimated by \texttt{systemfit}.
}
\note{
\begin{itemize}
\item Many theoretical models that are econometrically estimated
   consist of more than one equation.
\item For instance, economists like me model the supply
   and the demand of a certain good.
\item Or forest engineers like my co-author Jeff model the growth of trees
   and take the height as well as the diameter of the trees into account.
\item The disturbance terms of the single equations of such a equation system
   are likely contemporaneously correlated.
\item For instance, if you have a specific tree with a below average
   diameter growth,
   the growth of the height will probably be also below average.
\item A standard OLS regression of each equation separately
   ignores this correlation and leads to inefficient estimates.
\item However, efficients estimates can be obtained,
   if all equations are estimated simultaneously,
   using Zellner's °Seemingly Unrelated Regression° method.
\myhrule
\item Furthermore, when the underlying theory suggests cross-equation
   parameter restrictions,
   a simultaneous estimation of all equations required, too.
\myhrule
\item Finally, it might be that some variables are the endogenous variable
    in some equations and a regressor in other equations.
\item In this case, °Two-Stage Least Squares° or °Three-Stage Least Squares°
   regressions of the system have to be applied to obtain unbiased estimates.
\myhrule
\item All these models can be estimated by \texttt{systemfit}.
\end{itemize}
}

% ----------------- Outline ----------------------------
\subsection{Outline}
\frame{
   \frametitle{Outline}
\begin{itemize}
\setlength{\itemsep}{0.75em}
\item Introduction
\item Features of systemfit
\item Examples
% \item Advantages
\item Plans for the Future
\end{itemize}
}
\note{
\begin{itemize}
\item After this introduction, I will present the features of systemfit.
\item Then, I will demonstrate the usage of systemfit by a simple example.
% \item Thereafter, I will highlight some advantages of systemfit compared to
%    other software packages.
\item Finally, I will present our future plans to extend and improve systemfit.
\end{itemize}
}

@@

% ----------------- Estimation Methods ----------------------------
\subsection{Methods}
\frame{
   \frametitle{Estimation Methods}
\begin{itemize}
\setlength{\itemsep}{0.75em}
\item Ordinary Least Squares (OLS)
\item Two-Stage Least Squares (2SLS)
\item Seemingly Unrelated Regression (SUR)
\item Three-Stage Least Squares (3SLS)
\item \ldots
\end{itemize}
}
\note{
\begin{itemize}
\item The most important features of \texttt{systemfit}
   are -- of course -- the estimation methods.
\item It provides
   \begin{itemize}
   \item Ordinary Least Squares
   \item Two-Stage Least Squares
   \item Seemingly Unrelated Regression
   \item Three-Stage Least Squares
   \item and some other less important methods
   \end{itemize}
\item Although the °Two-Stage Least Squares° estimator is designed
   for estimating equation systems,
   it can be applied to single equations, too.
\item I have heard from several people -- among them Jeff Racine --
   that they generally use systemfit to fit single-equation models
   by °Two-Stage Least Squares°.
\end{itemize}
}

% ----------------- Estimation Control ----------------------------
\subsection{Estimation Control}
\frame{
   \frametitle{Estimation Control}
\begin{itemize}
\setlength{\itemsep}{0.75em}
\item imposition of linear restrictions
\item instrumental variables
\item iterated estimation
\item formulas for the residual covariance matrix
\item formulas for 3SLS estimation
% \item degrees of freedom for $t$ tests
% \item homogenous residual variance in OLS/2SLS estimations
\item \ldots
\end{itemize}
}
\note{
\begin{itemize}
\item Besides the estimation method,
   the user can control many details of the estimation.
\item For instance, she or he can impose linear restrictions
   by two different methods,
\item and can specify instrumental variables:
   either the same for all equations or different for each equation.
\item Furthermore, it can be indicated whether the feasable least squares
   estimations should be iterated.
\item The literature suggests several different formulas to calculate
   the residual covariance matrix and the Three-Stage Least Squares estimator.
\item To my knowledge, systemfit is the only popular software
   where the user can decide which formula is applied.
\item There exist some further details that can be controlled by the user.
\item If you are interested in them please don't hesitate to ask me or
   take a look into the documentation.
\end{itemize}
}

% ----------------- Other Tools ----------------------------
\subsection{Other Tools}
\frame{
   \frametitle{Other Tools}
\begin{itemize}
\setlength{\itemsep}{0.75em}
\item wrapper function for (classical) panel-like data in long format:
   \texttt{systemfitClassic}
\item testing linear hypotheses using the F-, Wald-, and LR-statistic
\item Hausman test for the consistency of the 3SLS estimator
\end{itemize}
}
\note{
\begin{itemize}
\item The package systemfit does not just provide the function systemfit
   to estimate equation systems,
   but also a few other functions.
\item For instance, the package includes the wrapper function systemfitClassic
   that can be applied to panel-like data in long format.
\item Then, there are functions to test linear hypotheses using the
   F-, Wald-, and Likelihood-ratio-statistic
\item Finally, the consistency of the 3SLS estimator can be tested
   by a hausman test.
\end{itemize}
}

@@

% ----------------- Commands ----------------------------
\subsection{Commands}
\begin{SaveVerbatim}{specSystem}
eqDemand <- consump ~ price + income
eqSupply <- consump ~ price + farmPrice + trend
system <- list(demand=eqDemand, supply=eqSupply)
\end{SaveVerbatim}
% \end{Verbatim}
\begin{SaveVerbatim}{estSystem}
fitsur <- systemfit("SUR", system, data=Kmenta)
\end{SaveVerbatim}
% \end{Verbatim}
\begin{SaveVerbatim}{printResults}
summary( fitsur )
\end{SaveVerbatim}
% \end{Verbatim}
\frame{
   \frametitle{Example: Commands}
\begin{itemize}
\setlength{\itemsep}{0.75em}
\item from Kmenta (1986): Elements of Econometrics, p.~685
\item specification of the equation system:\\[2mm]
   {\small \BUseVerbatim{specSystem}}
\item estimation using method °SUR°:
   {\small \BUseVerbatim{estSystem}}
\item printing summary results:\\
   {\small \BUseVerbatim{printResults}}
\end{itemize}
}
\note{
\begin{itemize}
\item ?
\end{itemize}
}

@@

% ============== Advantages ==============================
% \section{Advantages}
% \frame{
%    \frametitle{Advantages}
% \begin{itemize}
% \item all estimation procedures are publicly available
% \item estimation algorithms can be easily modified
% \item many estimation details can be controlled by the user
% \item novices can safely ignore these estimation details
% \item reliability of \texttt{systemfit} has been demonstrated
% \end{itemize}
% }
% \note{
% \begin{itemize}
% \item all estimation procedures are publicly available in the source code
% \item estimation algorithms can be easily modified to meet specific
% requirements
% \item the (advanced) user can control many estimation details generally
% not available in other software packages
% \item novices can safely ignore these estimation details
% because they have reasonable defaults
% \item reliability of \texttt{systemfit} has been demonstrated by comparing its estimation
% results with results published in the literature
% \end{itemize}
% }

@@

% ----------------- General Plans for the Future ----------------------------
\subsection{General}
\frame{
   \frametitle{Plans for the Future}
\begin{itemize}
\setlength{\itemsep}{0.75em}
\item estimation with unbalanced data sets
\item estimation methods: LIML, FIML,  and GMM
\item fitting equation systems with serially correlated and
heteroscedastic disturbances
\item spatial econometric methods
\item improving the function \texttt{nlsystemfit}
to estimate systems of non-linear estimations
\end{itemize}
}
\note{
\begin{itemize}
\item ?
\end{itemize}
}
